

# Gráficos com o **ggplot2**

Fonte: [http://tutorials.iq.harvard.edu](http://tutorials.iq.harvard.edu)

## Introdução

O pacote **ggplot2** ...

## Gramática dos gráficos

O pacote **ggplot2** ...

Existem outros pacotes gráficos em R, alguns independentes mas outros que estendem ainda mais a abrangência do pacote *ggplot2*, como por exemplo:

     - Gráficos 3-D (pacote rgl)
     - Plotagem de grafos (pacote *igraph*)
     - Gráficos interativos (pacote *ggvis*)



## Primeiro exemplo
 
### Dados: "*Housing prices*"


```{r}
#   Importe os dados (arquivo landdata-states-2016q1.xls).
housing <- read.csv("https://efsilvaa.github.io/IntroR/dados/landdata-states.csv")
head(housing[1:5])
```

### **ggplot2** VS Base 

Comparado com o pacote *base*, *ggplot2* :

       - comandos com mais texto para gráficos simples
       - porém mais comandos mais simples para gráficos complexos
       - dados sempre devem estar em um `data.frame`
       - usa uma sistemática diferente para adicionar os elementos gráficos
       

Como já vimos, para montar um histograma com o pacote *base*:

```{r}
hist(housing$Home.Value)
```

Já usando *ggplot2*:
```{r warning=FALSE}
library(ggplot2) #não esqueça de carregar e instalar o pacote 

ggplot(housing, aes(x = Home.Value)) +    
       geom_histogram()
```

*Obs:* '+' significa que o próximo comando será somado na definição do gráfico   

>  *Base* wins!

### *ggplot2* VS *base* agora com gráficos mais complexos:

O gráfico apresenta a evolução dos preços das casas em dois estados americanos, onde os pontos de cada estado possui uma cor distinta. 

```{r}
plot(Home.Value ~ Date,
     data=subset(housing, State == "MA"))

points(Home.Value ~ Date, col="red",
       data=subset(housing, State == "TX"))

legend(1975, 400000,
       c("MA", "TX"), title="State",
       col=c("black", "red"),
       pch=c(1, 1))

```


Já com *ggplot2*:

```{r}
ggplot(subset(housing, State %in% c("MA", "TX")),
       aes(x=Date,
           y=Home.Value,
           color=State))+
  
  geom_point()

```

>   *ggplot2* wins!

## Estrutura para montagem de um gráfico

```{r eval=FALSE}
ggplot(data = <default data set>, 
         aes(x = <default x axis variable>,
             y = <default y axis variable>,
             ... <other default aesthetic mappings>),
         ... <other plot defaults>) +

         geom_<geom type>(aes(size = <size variable for this geom>, 
                        ... <other aesthetic mappings>),
                    data = <data for this point geom>,
                    stat = <statistic string or function>,
                    position = <position string or function>,
                    color = <"fixed color specification">,
                    <other arguments, possibly passed to the _stat_ function) +

    scale_<aesthetic>_<type>(name = <"scale label">,
                       breaks = <where to put tick marks>,
                       labels = <labels for tick marks>,
                       ... <other options for the scale>) +

    theme(plot.background = element_rect(fill = "gray"),
          ... <other theme elements>)
```


## Mapeamento Visual e geometrias (Aesthetic Mapping and geometrics)

### Aesthetic Mapping

...

### Objetos geométricos (`geom`)


Uma lista das `geom_` disponíveis pode ser visualizada usando o comando `help.search("geom_", package = "ggplot2")` ou simplesmente teclando `geom_<tab>` em uma boa IDE R (como o Rstudio)


## Pontos (Scatterplot)

`geom_point' requer que pelo menos o mapeamento de x e y tenham sido feitos.  Os outros são opcionais.


```{r}
hp2001Q1 <- subset(housing, Date == 2001.25) 
ggplot(hp2001Q1,
       aes(y = Structure.Cost, x = Land.Value)) +
       geom_point()
```

Usando a `log()` para transformar a variável no eixo x.
```{r}
ggplot(hp2001Q1,
       aes(y = Structure.Cost, x = log(Land.Value))) +
       geom_point()

```



## Linhas (Ex. Regressão linear)

Como já foi dito, um gráfico usando `ggplot2` pode conter mais de uma `geom`. Nesse caso, funciona como novas camadas fossem acrescentadas ao gráfico.


```{r}
hp2001Q1$pred.SC <- predict(lm(Structure.Cost ~ log(Land.Value), data = hp2001Q1))

p1 <- ggplot(hp2001Q1, aes(x = log(Land.Value), y = Structure.Cost))

p1 + geom_point(aes(color = Home.Value)) +
     geom_line(aes(y = pred.SC))

```

## Smoothers

  Esse `geom_` inclui uma linha de predição e uma faixa de intervalo de confiança considerando os pontos informados em `aes()`.  Desta vez vamos reaproveitar nossa variável `p1`.

```{r}
p1 +
  geom_point(aes(color = Home.Value)) +
  geom_smooth()

```


## Texto (Label Points)

```{r}
p1 + 
  geom_text(aes(label=State), size = 3)

```
*obs:* Para mais possibilidades com textos existe o pacote *ggrepel* [@R-ggrepel].
```{r}
## install.packages("ggrepel") 
  library("ggrepel")
  p1 + 
  geom_point() + 
  geom_text_repel(aes(label=State), size = 3)
```


##  Aesthetic Mapping VS Assignment


Por exemplo: 
```{r}
p1 +
  geom_point(aes(size = 2),  # Errado! 2 não é uma variável
             color="red")    # ok -- todos os pontos devem ser *red*

```


Mais de uma variável mapeada em `aes()`:

```{r}
p1 +
  geom_point(aes(color=Home.Value, shape = region))
```


## Exercicio I


Os dados para este exercício estão em no arquivo 'EconomistData.csv'.

Esses dados revelam o Índice de Desenvolvimento Humano(HDI) e o Índice Corrupção Perceptível(CPI) de diversos países.
    
```{r eval= FALSE}
dat <- read.csv(""https://efsilvaa.github.io/IntroR/dados/EconomistData.csv")
head(dat)
```

Se tudo estive correto, o comando abaixo deve funcionar:
```{r eval = FALSE, include = FALSE}
ggplot(dat, aes(x = CPI, y = HDI, size = HDI.Rank))
     + geom_point()

```


    1. Crie um gráfico de pontos (scatter plot) com CPI no eixo x e HDI no eixo y.
    2. Colorir os pontos em azul (blue).
    3. Use a região para mapear a cor dos pontos.
    4. Aumente o tamanho dos pontos (size = 2)
    5. Associe o tamanho dos pontos ao HDI.Rank


```{r eval= FALSE, include= FALSE}

## Statistical Transformations
## ═══════════════════════════

## Statistical Transformations
## ───────────────────────────

##   Some plot types (such as scatterplots) do not require
##   transformations–each point is plotted at x and y coordinates equal to
##   the original value. Other plots, such as boxplots, histograms,
##   prediction lines etc. require statistical transformations:
##   • for a boxplot the y values must be transformed to the median and
##     1.5(IQR)
##   • for a smoother smother the y values must be transformed into
##     predicted values

##   Each `geom' has a default statistic, but these can be changed. For
##   example, the default statistic for `geom_bar' is `stat_bin':
args(geom_histogram)
args(stat_bin)

## Setting Statistical Transformation Arguments
## ────────────────────────────────────────────

##   Arguments to `stat_' functions can be passed through `geom_'
##   functions. This can be slightly annoying because in order to change it
##   you have to first determine which stat the geom uses, then determine
##   the arguments to that stat.

##   For example, here is the default histogram of Home.Value:
p2 <- ggplot(housing, aes(x = Home.Value))
p2 + geom_histogram()

##   The binwidth looks reasonable by default, but we can change it by
##   passing the `binwidth' argument to the `stat_bin' function:
p2 + geom_histogram(stat = "bin", binwidth=4000)

## Changing The Statistical Transformation
## ───────────────────────────────────────

##   Sometimes the default statistical transformation is not what you need.
##   This is often the case with pre-summarized data:
housing.sum <- aggregate(housing["Home.Value"], housing["State"], FUN=mean)
rbind(head(housing.sum), tail(housing.sum))

ggplot(housing.sum, aes(x=State, y=Home.Value)) + 
  geom_bar()

##   What is the problem with the previous plot? Basically we take binned
##   and summarized data and ask ggplot to bin and summarize it again
##   (remember, `geom_bar' defaults to `stat = stat_count'); obviously this
##   will not work. We can fix it by telling `geom_bar' to use a different
##   statistical transformation function:
ggplot(housing.sum, aes(x=State, y=Home.Value)) + 
  geom_bar(stat="identity")

## Exercise II
## ───────────

##   1. Re-create a scatter plot with CPI on the x axis and HDI on the y
##      axis (as you did in the previous exercise).
##   2. Overlay a smoothing line on top of the scatter plot using
##      geom_smooth.
##   3. Overlay a smoothing line on top of the scatter plot using
##      geom_smooth, but use a linear model for the predictions. Hint: see
##      `?stat_smooth'.
##   4. Overlay a smoothling line on top of the scatter plot using
##      geom_line. Hint: change the statistical transformation.
##   5. BONUS: Overlay a smoothing line on top of the scatter plot using
##      the default /loess/ method, but make it less smooth. Hint: see
##      `?loess'.

## Scales
## ══════

## Scales: Controlling Aesthetic Mapping
## ─────────────────────────────────────

##   Aesthetic mapping (i.e., with `aes()') only says that a variable
##   should be mapped to an aesthetic. It doesn't say /how/ that should
##   happen. For example, when mapping a variable to /shape/ with
##   `aes(shape = x)' you don't say /what/ shapes should be used.
##   Similarly, `aes(color = z)' doesn't say /what/ colors should be used.
##   Describing what colors/shapes/sizes etc. to use is done by modifying
##   the corresponding /scale/. In `ggplot2' scales include
##   • position
##   • color and fill
##   • size
##   • shape
##   • line type
##   Scales are modified with a series of functions using a
##   `scale_<aesthetic>_<type>' naming scheme. Try typing `scale_<tab>' to
##   see a list of scale modification functions.

## Common Scale Arguments
## ──────────────────────

##   The following arguments are common to most scales in ggplot2:
##   name: the first argument gives the axis or legend title
##   limits: the minimum and maximum of the scale
##   breaks: the points along the scale where labels should appear
##   labels: the labels that appear at each break

##   Specific scale functions may have additional arguments; for example,
##   the `scale_color_continuous' function has arguments `low' and `high'
##   for setting the colors at the low and high end of the scale.

## Scale Modification Examples
## ───────────────────────────

##   Start by constructing a dotplot showing the distribution of home
##   values by Date and State.

p3 <- ggplot(housing,
             aes(x = State,
                 y = Home.Price.Index)) + 
        theme(legend.position="top",
              axis.text=element_text(size = 6))
(p4 <- p3 + geom_point(aes(color = Date),
                       alpha = 0.5,
                       size = 1.5,
                       position = position_jitter(width = 0.25, height = 0)))

##   Now modify the breaks for the x axis and color scales

p4 + scale_x_discrete(name="State Abbreviation") +
  scale_color_continuous(name="",
                         breaks = c(1976, 1994, 2013),
                         labels = c("'76", "'94", "'13"))

##   Next change the low and high values to blue and red:
p4 +
  scale_x_discrete(name="State Abbreviation") +
  scale_color_continuous(name="",
                         breaks = c(1976, 1994, 2013),
                         labels = c("'76", "'94", "'13"),
                         low = "blue", high = "red")

p4 +
  scale_color_continuous(name="",
                         breaks = c(1976, 1994, 2013),
                         labels = c("'76", "'94", "'13"),
                         low = muted("blue"), high = muted("red"))

## Using different color scales
## ────────────────────────────

##   ggplot2 has a wide variety of color scales; here is an example using
##   `scale_color_gradient2' to interpolate between three different colors.
p4 +
  scale_color_gradient2(name="",
                        breaks = c(1976, 1994, 2013),
                        labels = c("'76", "'94", "'13"),
                        low = muted("blue"),
                        high = muted("red"),
                        mid = "gray60",
                        midpoint = 1994)

## Available Scales
## ────────────────

##   • Partial combination matrix of available scales
##   ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
##    *Scale*          *Types*     *Examples*              
##   ──────────────────────────────────────────────────────
##    scale_color_     identity    scale_fill_continuous   
##    scale_fill_      manual      scale_color_discrete    
##    scale_size_      continuous  scale_size_manual       
##                     discrete    scale_size_discrete     
                                                      
##    scale_shape_     discrete    scale_shape_discrete    
##    scale_linetype_  identity    scale_shape_manual      
##                     manual      scale_linetype_discrete 
                                                      
##    scale_x_         continuous  scale_x_continuous      
##    scale_y_         discrete    scale_y_discrete        
##                     reverse     scale_x_log             
##                     log         scale_y_reverse         
##                     date        scale_x_date            
##                     datetime    scale_y_datetime        
##   ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

##   Note that in RStudio you can type `scale_' followed by TAB to get the
##   whole list of available scales.

## Exercise III
## ────────────

##   1. Create a scatter plot with CPI on the x axis and HDI on the y axis.
##      Color the points to indicate region.
##   2. Modify the x, y, and color scales so that they have more
##      easily-understood names (e.g., spell out "Human development Index"
##      instead of "HDI").
##   3. Modify the color scale to use specific values of your choosing.
##      Hint: see `?scale_color_manual'.

## Faceting
## ════════

## Faceting
## ────────

##   • Faceting is `ggplot2' parlance for *small multiples*
##   • The idea is to create separate graphs for subsets of data
##   • `ggplot2' offers two functions for creating small multiples:
##     1. `facet_wrap()': define subsets as the levels of a single grouping
##        variable
##     2. `facet_grid()': define subsets as the crossing of two grouping
##        variables
##   • Facilitates comparison among plots, not just of geoms within a plot

## What is the trend in housing prices in each state?
## ──────────────────────────────────────────────────

##   • Start by using a technique we already know–map State to color:
p5 <- ggplot(housing, aes(x = Date, y = Home.Value))
p5 + geom_line(aes(color = State))

##   There are two problems here–there are too many states to distinguish
##   each one by color, and the lines obscure one another.

## Faceting to the rescue
## ──────────────────────

##   We can remedy the deficiencies of the previous plot by faceting by
##   state rather than mapping state to color.

(p5 <- p5 + geom_line() +
   facet_wrap(~State, ncol = 10))

##   There is also a `facet_grid()' function for faceting in two
##   dimensions.

## Themes
## ══════

## Themes
## ──────

##   The `ggplot2' theme system handles non-data plot elements such as
##   • Axis labels
##   • Plot background
##   • Facet label backround
##   • Legend appearance
##   Built-in themes include:
##   • `theme_gray()' (default)
##   • `theme_bw()'
##   • `theme_classc()'
p5 + theme_linedraw()

p5 + theme_light()

## Overriding theme defaults
## ─────────────────────────

##   Specific theme elements can be overridden using `theme()'. For
##   example:
p5 + theme_minimal() +
  theme(text = element_text(color = "turquoise"))

##   All theme options are documented in `?theme'.

## Creating and saving new themes
## ──────────────────────────────

##   You can create new themes, as in the following example:

theme_new <- theme_bw() +
  theme(plot.background = element_rect(size = 1, color = "blue", fill = "black"),
        text=element_text(size = 12, family = "Serif", color = "ivory"),
        axis.text.y = element_text(colour = "purple"),
        axis.text.x = element_text(colour = "red"),
        panel.background = element_rect(fill = "pink"),
        strip.background = element_rect(fill = muted("orange")))

p5 + theme_new

## The #1 FAQ
## ══════════

## Map Aesthetic To Different Columns
## ──────────────────────────────────

##   The most frequently asked question goes something like this: /I have
##   two variables in my data.frame, and I'd like to plot them as separate
##   points, with different color depending on which variable it is. How do
##   I do that?/

## Wrong
## ╌╌╌╌╌

housing.byyear <- aggregate(cbind(Home.Value, Land.Value) ~ Date, data = housing, mean)
ggplot(housing.byyear,
       aes(x=Date)) +
  geom_line(aes(y=Home.Value), color="red") +
  geom_line(aes(y=Land.Value), color="blue")


#

## Right
## ╌╌╌╌╌

library(tidyr)
home.land.byyear <- gather(housing.byyear,
                           value = "value",
                           key = "type",
                           Home.Value, Land.Value)
ggplot(home.land.byyear,
       aes(x=Date,
           y=value,
           color=type)) +
  geom_line()

## Putting It All Together
## ═══════════════════════

## Challenge: Recreate This `Economist' Graph
## ──────────────────────────────────────────

##   [file:images/Economist1.pdf]

##   Graph source: [http://www.economist.com/node/21541178]

##   Building off of the graphics you created in the previous exercises,
##   put the finishing touches to make it as close as possible to the
##   original economist graph.

##   [file:images/Economist1.png]

## Wrap-up
## ═══════


## Help Us Make This Workshop Better!
## ──────────────────────────────────

##   • Please take a moment to fill out a very short feedback form
##   • These workshops exist for you – tell us what you need!
##   • [http://tinyurl.com/R-graphics-feedback]

```


## Recursos adicionais 

ggplot2:

      - Mailing list: [http://groups.google.com/group/ggplot2](http://groups.google.com/group/ggplot2)
      - Wiki: [https://github.com/hadley/ggplot2/wiki](https://github.com/hadley/ggplot2/wiki)
      - Website: [http://had.co.nz/ggplot2/](http://had.co.nz/ggplot2/)
      - StackOverflow: [http://stackoverflow.com/questions/tagged/ggplot](http://stackoverflow.com/questions/tagged/ggplot)
